\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-11-21 smurf_paper.tex
more edits... figure panel labels, chi^2, acknowledge Dan...

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro}
\includegraphics[width=0.49\linewidth]{bolos_point_mix}
\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  Left: data taken during the S2SRO period,
  observation 29 on 20100313. There is a strong correlation between
  the bolometers and the roughly $\sim$25\,s oscillation in the
  fridge, but only a minor correlation with the opacity and telescope
  motion. The nearly flat common-mode subtracted signals show: (i)
  that most of the low-frequency signal is common to all of the
  bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. Right: data taken with the fully-commissioned
  instrument, observation 38 on 20111112. Unlike the S2SRO data, there
  is no strong signal produced by variations in the fridge
  temperature. The 450\,\micron\ data have minimal correlated
  low-frequency noise (as evidenced by the lack of a difference
  between the raw and common-mode subtracted signals), although the
  850\,\micron\ data show a common signal correlated with the opacity
  and telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set), mostly likely caused by a
  combination of changing airmass (anti-correlated with opacity), and
  magnetic field pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro}
\includegraphics[width=0.49\linewidth]{pspec}
\caption{Bolometer power spectral densities for the same two data sets
  (before and after upgrades) used in Fig.~\ref{fig:bolos_mix}. The
  PSDs have been boxcar smoothed with a width of 0.1\,Hz to reduce the
  noise slightly and clarify some features. Four of the most sensitive
  bolometers have been selected at each wavelength, and the
  flat-fielded and step-corrected (but otherwise raw) PSDs are shown
  as coloured dotted lines (the blue signals are for the same time
  series as those shown in Fig.~\ref{fig:bolos_mix}).  The solid black
  lines are the PSDs of the common-mode signals at each wavelength,
  and the solid coloured lines show the PSDs of the bolometers once
  the common-mode is removed.  Finally, the dashed black lines show
  the spectral shape produced by a point source given the scan speeds
  for the two observations (120\,arcsec\,s$^{-1}$ in the left-hand
  plot, and 190\,arcsec\,s$^{-1}$ in the right-hand plot); for
  reference, the top horizontal axes shows the conversion from
  frequency to angular scale. Horizontal dotted lines at $2 \times
  10^5$ and $1 \times 10^5$\,pW\,Hz$^{-1}$ at 450 and 850\,\micron,
  respectively, are provided as a visual reference for the white-noise
  levels. In addition to $1/f$ and white noise components, and line
  features, all of the PSDs exhibit the gradual roll-off of the
  anti-aliasing filter just below the Nyquist frequency.  Left: for
  the S2SRO data there are broad line features in the PSDs at both
  wavelengths above $\sim$35\,Hz.  At lower frequencies, the bolometer
  signals exhibit clear $1/f$ knees at approximately 1 and 2\,Hz at
  450 and 850\,\micron. Common-mode subtraction removes most of the
  correlated fridge oscillation signal, lowering the $1/f$ knees to
  approximately 0.2 and 0.7\,Hz at 450 and 850\,\micron,
  respectively. Right: data from the fully-commissioned instrument
  tend to have lower $1/f$ knees, and fewer line features. The
  remaining low-frequency noise, now that the fridge oscillations have
  been removed, is considerably more independent from bolometer to
  bolometer than in the S2SRO data (note the larger spread in the
  dotted lines at low frequencies compared to the left-hand plot,
  particularly at 450\,\micron). Finally, this example shows that the
  white noise performance is in fact similar for the two subarrays
  (s4a and s8b) that were used both before and after the upgrades.}
\label{fig:pspec}
\end{figure*}

@@

In the S2SRO data (left panel), observation 29 on 20100313, both
bolometers share significant long-timescale structure ($\gsim10$\,s)
that appears to be related to variations in the fridge base
temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, are comparable to those at
850\,\micron. Such behaviour might be expected if there is a
comparable varying thermal load from the fridge at each wavelength
that dominates. We also note that there is no obvious strong
correlation between the low-frequency signal structure, at either
wavelength, with the telescope motion.

@@

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in the right panel of
Fig.~\ref{fig:bolos_mix}. Having solved the fridge oscillation
problem, these data no longer exhibit a correlation with the mixing
chamber (however, note that variations are still seen in the mixing
chamber signal; such variations do not necessarily reflect changes in
the focal plane temperature). There is minimal correlated
low-frequency noise in these 450\,\micron\ data (note that there is
almost no difference once the common-mode has been subtracted). The
850\,\micron\ channel, however, exhibits a more significant signal
that is obviously correlated with the telescope motion. It is also
correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. Most of this ``scan-synchronous'' noise is
attributed to magnetic field pickup, as described in
Section~\ref{sec:magpickup}.

@@

The dotted coloured lines in Fig.~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded and step-corrected (Section~\ref{sec:steps})
data. At each wavelength, and in both data sets, there are clear $1/f$
knees at frequencies ranging from roughly 1 to 2\,Hz, followed by a
predominantly white spectrum punctuated by line features at higher
frequencies, and finally a roll-off caused by the anti-aliasing filter
above $\gsim 70$\,Hz. The correlation between the low-frequency
components of the different bolometer signals is large. The solid
black lines in Fig.~\ref{fig:pspec} indicate the PSDs of the
common-modes $\mathbf{c}(t)$ at each wavelength, which reproduce much
of the low-frequency structure, as well as some of the
higher-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high-frequency,
as expected if the bolometers are dominated by uncorrelated white
noise in that part of the spectrum. The common-mode signals are fit to
each bolometer time series and removed as in Fig.~\ref{fig:bolos_mix},
and the resulting PSDs are shown as solid coloured lines.  For
reference, the top horizontal axes have been converted to angular
scale assuming the scan speeds of each observation
(120\,arcsec\,s$^{-1}$ in the left-hand plot, and
190\,arcsec\,s$^{-1}$ in the right-hand plot). The power spectra of
unresolved point-sources are also shown as dashed lines in each band
(arbitrarily normalised), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features at these scan speeds (this may not be the
case at higher scan speeds).

In the S2SRO data (left-hand panel of Fig.~\ref{fig:pspec}), the
low-frequency noise is very correlated between the detectors (note the
tight scatter in the dotted coloured lines, and their resemblance to the
common-mode), due to it being dominated by the fridge
oscillations. Common-mode subtraction is very effective, reducing the
$1/f$ knees in both wavelengths by about a factor of 5.

Raw data from the fully-commissioned instrument (right-hand panel of
Fig.~\ref{fig:pspec}) generally have less significant $1/f$ noise as
compared with the S2SRO data, and it is considerably less correlated
amongst detectors (larger spread in the dotted coloured lines,
particularly noticeable in these 450\,\micron\ data), leading to a
less drastic improvement upon common-mode removal. However, since the
data are less dominated by low-frequency noise to begin with, the
signals generally require less aggressive high-pass filtering to
produce maps, and therefore retain larger-scale structures than with
the S2SRO data. Furthermore, there are generally fewer line features
in the PSDs of bolometers in the fully-commissioned
instrument. Finally, these two data sets illustrate that the white
noise performance (NEFD) is in fact similar before and after the
upgrades for these two subarrays (s4a at 450\,micron, and s8b at
850\,\micron). The main improvements are a reduction in correlated and
line noise features mentioned above, the larger number of working
bolometers and field-of-view (approximately a factor of 4).

@@

A novel feature of our \scuba\ analysis is that we can perform PCA
with 450 and 850\,\micron\ data simultaneously, potentially helping us
to differentiate thermal and optical noise signals (e.g., from the
atmosphere) that might appear in both wavelengths, from other noise
sources that are restricted to single subarrays (such as readout
noise). In Fig.~\ref{fig:pca} we show the results for a combined
analysis of the s4a and s8b subarrays for the first $\sim$100\,s of
the same observation that was used in the right-hand
(fully-commissioned) panels of Figs.~\ref{fig:bolos_mix} and
\ref{fig:pspec}. The 12 most significant components are shown (ranked
by the mean eigenvalues across all working bolometers in both
bands). The top and middle panels for each component show the
time-series and PSDs of the normalised eigenvectors. The bottom panels
are maps indicating the eigenvalues (amplitudes) of the component
across the focal plane (with the mean $\bar{\lambda}$ and and \rms,
$\lambda_\mathrm{rms}$, of the eigenvalues for all of the bolometers
calculated separately at each wavelength also shown).

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is clearly produced by Components~1 and 2. In both
cases, the eigenvector time-series exhibit a roughly periodic signal
that resembles the scan pattern in the right-hand panel of
Fig.~\ref{fig:bolos_mix}. While these correlated signals also clearly
appear at both wavelengths, referring to the maps of eigenvalues, they
are considerably stronger at 850\,\micron\ (consistent with the visual
appearance of the bolometer signals in Fig.~\ref{fig:pca}). Also note
that the PSDs for these eigenvectors exhibit nearly a pure $1/f$
signature with almost no line features, or a white-noise plateau
(suggesting that these are high-\snr\ measurements of a purely
low-frequency drift).

\begin{figure*}
\centering
\includegraphics[width=0.85\linewidth]{pca}
\caption{The first nine components from a principal component
  analysis, ranked by the mean eigenvalues of combined 450 (s4a
  subarray) and 850\,\micron\ (s8b subarray) time-series bolometer
  data, for the same observation as that used in the right-hand panels
  of Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec}. For each component
  the top plot shows the time-series of the eigenvectors normalised by
  their \rms, the middle plot its PSD, and the bottom coloured panels
  the eigenvalues for the bolometers at each wavelength (the
  amplitudes of the eigenvectors in the time-series). For reference,
  both the mean, $\bar{\lambda}$, and \rms, $\lambda_\mathrm{rms}$,
  eigenvalues for the bolometers in each subarray are also shown. Most
  of the correlated $1/f$ signal is encapsulated in Component~1, and
  appears to be dominated by scan-synchronous noise (compare with the
  scan pattern in the right-hand panel of
  Fig.~\ref{fig:bolos_mix}). Some of this signal also appears in
  Component~3. Component~2 is another $1/f$ signal that is dominant in
  the 850\,\micron\ subarray. Most of the remaining components consist
  of weaker $1/f$ noise and line features predominantly above
  $\sim$30\,Hz. The eigenvalue maps show correlated patterns along
  columns suggesting pickup in their common amplifier chains.}
\label{fig:pca}
\end{figure*}

@@

The approach taken by SMURF to reducing \scuba\ data is to model (and
remove) predominantly low-frequency noise sources that are correlated
amongst detectors, and to iterate this process along with estimates of
the map. Significant experimentation with different models for noise
sources during \scuba\ commissioning required a highly flexible
software framework, and a configurable user interface. To achieve this
goal, while minimizing development time, it was decided to build SMURF
as a Starlink package \citep{2009ASPC..411..418J}, which provides
access to a large suite of libraries (including a commanding and
messaging interface, astrometric coordinate conversions and generation
of standard WCS information, file formats etc.). Furthermore,
Starlink\footnote{\url{http://www.starlink.ac.uk}} is
open-source\footnote{\url{https://github.com/Starlink}} (distributed
under the GNU General Public Licence v3), and already used extensively
at the Joint Astronomy Centre (host of the JCMT) for many other
systems \citep{jenness2011}, which helps with interoperability. Though
originally written in FORTRAN, many of the core Starlink libraries are
now ported to native C, or at least have a C interface. It was
therefore decided to develop SMURF in C as well (rather than C++, for
example, which would have required adding further dependencies to
Starlink), although we have taken an object-oriented approach. For
example, all of the data for a given noise model are encapsulated in a
C structure, and there is a standard interface for all functions that
handle models (member data and functions for the class,
respectively). In this way it is easy to extend SMURF with new
models. Parallelisation is incorporated in the most time-consuming
low-level routines using threads (e.g., performing Fast Fourier
Transforms, re-gridding the data), usually handling either independent
blocks of bolometers, or blocks of time, in each thread, depending on
the nature of the calculation. We have found that for typical data
sets the processing time scales well with the number of central
processing unit (CPU) cores, although beyond 8 the returns are
diminished due to tasks that cannot be parallelised (e.g., reading the
data from disk).

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{steps1}
\caption{Examples of steps in time-series data. Steps occur with a wide
range of heights from the large steps shown on the left to the small
steps shown on the right. Large steps are often followed by a brief
``over-shoot'' as shown on the left. In both plots, the black curve is
the uncorrected time-series, and the red curve is the corrected time
series. Samples close to a step are omitted in the corrected time-series.
In the right hand plot, the blue curve is the uncorrected time-series for
a nearby bolometer. The similarity between the red and blue curves shows
that the step correction is performing well.
}
\label{fig:steps1}
\end{figure*}

@@

Steps occur with a wide range of heights and shapes. The ratio of step
height to noise can vary from less than 10 to several hundred. Some
steps occur over a single sample, such as the step close to sample
5000 in the right-hand plot of Fig.~\ref{fig:steps1}, but others
happen more gradually, such as the step close to sample 5300. In
addition, a step can be preceded or followed by a short period of
instability, as is visible at the bottom of the step in the left hand
plot of Fig.~\ref{fig:steps1} (this is probably due to the response of
the \scuba\ anti-aliasing filter; the sudden large step occurs prior
to the 200\,Hz re-sampling). Further problems are caused by steps that
occur close together in time, such as the large downward step followed
by a smaller upward step close to sample 5000 in the right hand plot
of Fig.~\ref{fig:steps1}.

@@

The primary purpose of the noise component, \model{NOI}, is to measure
the white noise levels of each bolometer [$\mathbf{n}^\mathrm{w}_i(t)$
in Eq.~\ref{eq:noise}]. The measurement generally occurs once all of
the other models have been fit and removed. This noise may then be
used to estimate weights for the data in subsequent iterations.

@@

The map-maker will halt if the convergence criteria have been
met. Presently there are two numerical tests that may be performed by
SMURF, or alternatively, a fixed number of iterations may be
specified. The first numerical test is an approximate check of the
change in reduced chi-squared, $\chi^2_\mathrm{r}$. The standard
deviations, $\sigma_i$, are measured for each of the residual
bolometer signals once the modelled signal components have been
removed. In the early iterations, the residual signals usually contain
both white noise, and other long-timescale features. However, once the
solution has converged, this signal should look approximately
white. Once the white noise, $\mathbf{n}^\mathrm{w}_i(t)$, has been
measured from the bolometer PSDs in \model{NOI}
(Section~\ref{sec:noi}), $\chi^2_\mathrm{r}$ is calculated as $(1/N)
\sum_i \sigma^2_i / \mathbf{n}^\mathrm{w}_i(t)$, where $i$ runs over
the $N$ bolometers. In other words, it is the average ratio between
the measured time-series bolometer variances and their white-noise
levels measured between 2 and 10\,Hz. Note that this expression does
not account for the degrees of freedom. Clearly there are a large
number of parameters in the model, which will ``fit-out'' some of the
uncorrelated white noise, and potentially bias this estimate of
$\chi^2_\mathrm{r}$ low. One approach would be to keep track of the
degrees of freedom, and therefore provide a correction, as in
\citet{kovacs2008}. However, note that we are only using this quantity
as a convergence test, and its value is not important in an absolute
sense. Furthermore, the reference white noise values are calculated
only once in SMURF (Section~\ref{sec:noi}), and the number of model
parameters are fixed, so the value of $\chi^2_\mathrm{r}$ should
decrease monotonically if the solution is behaving well. The
convergence criterion is met if the change in subsequent values of
$\chi^2_\mathrm{r}$ is smaller than the requsted threshold, which is
set to $10^{-3}$ by default.

It was immediately found in the analysis of \scuba\ data that the
$\chi^2_\mathrm{r}$ test described above is not a sufficient stopping
criterion. The reason for this is that various components of the model
are, under normal circumstances, highly degenerate. One simple example
is the degeneracy between large-scale astronomical structures (larger
than the array footprint), and the common-mode rejection step
(\model{COM}), which will be illustrated in
Section~\ref{sec:point}. Since the flatness of the residual only
reflects the ability of the model to fit the shape of the bolometer
time-series, it does not necessarily correlate with convergence of the
most important model component: the map. Indeed, examination of the
map after large numbers of iterations (e.g., $>10$), shows the
presence of large-scale structures that are anti-correlated with
structures in other model components. For this reason, a second
map-based convergence test was added.

>>>>

\begin{figure*}
\centering

\begin{minipage}[h]{0.495\linewidth}
\textbf{(a) S2SRO} \\

\includegraphics[width=\linewidth]{bolos_point_mix_s2sro}
\end{minipage}
\begin{minipage}[h]{0.495\linewidth}
\textbf{(b) fully-commissioned} \\

\includegraphics[width=\linewidth]{bolos_point_mix}
\end{minipage}

\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  (a) Data taken during the S2SRO period,
  observation 29 on 20100313. There is a strong correlation between
  the bolometers and the roughly $\sim$25\,s oscillation in the
  fridge, but only a minor correlation with the opacity and telescope
  motion. The nearly flat common-mode subtracted signals show: (i)
  that most of the low-frequency signal is common to all of the
  bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. (b) Data taken with the fully-commissioned instrument,
  observation 38 on 20111112. Unlike the S2SRO data, there is no
  strong signal produced by variations in the fridge temperature. The
  450\,\micron\ data have minimal correlated low-frequency noise (as
  evidenced by the lack of a difference between the raw and
  common-mode subtracted signals), although the 850\,\micron\ data
  show a common signal correlated with the opacity and telescope
  motion (note that this scan has a significantly larger amplitude
  than the S2SRO data set), mostly likely caused by a combination of
  changing airmass (anti-correlated with opacity), and magnetic field
  pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

\begin{figure*}
\centering

\begin{minipage}[h]{0.495\linewidth}
\textbf{(a) S2SRO} \\

\includegraphics[width=\linewidth]{pspec_s2sro}
\end{minipage}
\begin{minipage}[h]{0.495\linewidth}
\textbf{(b) fully-commissioned} \\

\includegraphics[width=\linewidth]{pspec}
\end{minipage}

\caption{Bolometer power spectral densities (PSDs) for the same two
  data sets (before and after upgrades) used in
  Fig.~\ref{fig:bolos_mix}. The PSDs have been boxcar smoothed with a
  width of 0.1\,Hz to reduce the noise slightly and clarify some
  features. Four of the most sensitive bolometers have been selected
  at each wavelength, and the flat-fielded and step-corrected (but
  otherwise raw) PSDs are shown as coloured dotted lines (the blue
  signals are for the same time series as those shown in
  Fig.~\ref{fig:bolos_mix}).  The solid black lines are the PSDs of
  the common-mode signals at each wavelength, and the solid coloured
  lines show the PSDs of the bolometers once the common-mode is
  removed.  Finally, the dashed black lines show the spectral shape
  produced by a point source given the scan speeds for the two
  observations (120\,arcsec\,s$^{-1}$ in (a), and
  190\,arcsec\,s$^{-1}$ in (b); for reference, the top horizontal axes
  shows the conversion from frequency to angular scale. Horizontal
  dotted lines at $10^{-7}$ and $10^{-8}$\,pW\,Hz$^{-1}$ at 450 and
  850\,\micron, respectively, are provided as a visual reference for
  the white-noise levels. In addition to $1/f$ and white noise
  components, and line features, all of the PSDs exhibit the gradual
  roll-off of the anti-aliasing filter just below the Nyquist
  frequency.  (a) For the S2SRO data, particularly at 450\,\micron,
  there are broad line features in the PSDs at both wavelengths above
  $\sim$35\,Hz.  At lower frequencies, the bolometer signals exhibit
  clear $1/f$ knees at approximately 2 and 3\,Hz at 450 and
  850\,\micron. Common-mode subtraction removes most of the correlated
  fridge oscillation signal, lowering the $1/f$ knees to approximately
  0.6 and 1.0\,Hz at 450 and 850\,\micron, respectively. (b) Data from
  the fully-commissioned instrument tend to have lower $1/f$ knees,
  and fewer line features. For this example, the improvement is most
  striking at 450\,micron\ where the PSD is approximately two
  orders-of-magnitude lower in the fully-commissioned data at
  0.1\,Hz. The white noise performance, however, is similar for the
  two subarrays (s4a and s8b) that were used both before and after the
  upgrades.}
\label{fig:pspec}
\end{figure*}

@@

In the S2SRO data (Fig.~\ref{fig:bolos_mix}a), observation 29 on
20100313, both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, are comparable to those at
850\,\micron. Such behaviour might be expected if there is a
comparable varying thermal load from the fridge at each wavelength
that dominates. We also note that there is no obvious strong
correlation between the low-frequency signal structure, at either
wavelength, with the telescope motion.

@@

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data (note that
there is almost no difference once the common-mode has been
subtracted). The 850\,\micron\ channel, however, exhibits a more
significant signal that is obviously correlated with the telescope
motion. It is also correlated amongst many of the detectors, and
common-mode removal corrects it to a large extent. Part of the reason
that this is seen here, and not in the S2SRO data set shown, is that
the amplitude of the scan pattern is larger. Most of this
``scan-synchronous'' noise is attributed to magnetic field pickup, as
described in Section~\ref{sec:magpickup}.

@@

The dotted coloured lines in Fig.~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded and step-corrected (Section~\ref{sec:steps})
data. At each wavelength, and in both data sets, there are clear $1/f$
knees at frequencies ranging from roughly 1 to 2\,Hz, followed by a
predominantly white spectrum punctuated by line features at higher
frequencies, and finally a roll-off caused by the anti-aliasing filter
above $\gsim 70$\,Hz. The correlation between the low-frequency
components of the different bolometer signals is large. The solid
black lines in Fig.~\ref{fig:pspec} indicate the PSDs of the
common-modes $\mathbf{c}(t)$ at each wavelength, which reproduce much
of the low-frequency structure, as well as some of the
higher-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high-frequency,
as expected if the bolometers are dominated by uncorrelated white
noise in that part of the spectrum. The common-mode signals are fit to
each bolometer time series and removed as in Fig.~\ref{fig:bolos_mix},
and the resulting PSDs are shown as solid coloured lines.  For
reference, the top horizontal axes have been converted to angular
scale assuming the scan speeds of each observation
(120\,arcsec\,s$^{-1}$ in Fig.~\ref{fig:pspec}a, and
190\,arcsec\,s$^{-1}$ in Fig.~\ref{fig:pspec}b). The power spectra of
unresolved point-sources are also shown as dashed lines in each band
(arbitrarily normalised), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features at these scan speeds (this may not be the
case at higher scan speeds).

In the S2SRO data (Fig.~\ref{fig:pspec}a), the low-frequency noise is
very correlated between the detectors (note the tight scatter in the
dotted coloured lines, and their resemblance to the common-mode), due
to it being dominated by the fridge oscillations. Common-mode
subtraction is very effective, reducing the $1/f$ knees in both
wavelengths by about a factor of 5.

Raw data from the fully-commissioned instrument
(Fig.~\ref{fig:pspec}b) generally have less significant $1/f$ noise as
compared with the S2SRO data, and it is considerably less correlated
amongst detectors (larger spread in the dotted coloured lines,
particularly noticeable in these 450\,\micron\ data), leading to a
less drastic improvement upon common-mode removal. However, since the
data are less dominated by low-frequency noise to begin with, the
signals generally require less aggressive high-pass filtering to
produce maps, and therefore retain larger-scale structures than with
the S2SRO data. Furthermore, there are generally fewer line features
in the PSDs of bolometers in the fully-commissioned
instrument. Finally, these two data sets illustrate that the white
noise performance (NEFD) is in fact similar before and after the
upgrades for these two subarrays (s4a at 450\,micron, and s8b at
850\,\micron). The main improvements are a reduction in correlated and
line noise features mentioned above, the larger number of working
bolometers and field-of-view (approximately a factor of 4).

@@

A novel feature of our \scuba\ analysis is that we can perform PCA
with 450 and 850\,\micron\ data simultaneously, potentially helping us
to differentiate thermal and optical noise signals (e.g., from the
atmosphere) that might appear in both wavelengths, from other noise
sources that are restricted to single subarrays (such as readout
noise). In Fig.~\ref{fig:pca} we show the results for a combined
analysis of the s4a and s8b subarrays for the first $\sim$100\,s of
the same observation that was used in Figs.~\ref{fig:bolos_mix}b and
\ref{fig:pspec}b. The 12 most significant components are shown (ranked
by the mean eigenvalues across all working bolometers in both
bands). The top and middle panels for each component show the
time-series and PSDs of the normalised eigenvectors. The bottom panels
are maps indicating the eigenvalues (amplitudes) of the component
across the focal plane (with the mean $\bar{\lambda}$ and and \rms,
$\lambda_\mathrm{rms}$, of the eigenvalues for all of the bolometers
calculated separately at each wavelength also shown).

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is clearly produced by Components~1 and 2. In both
cases, the eigenvector time-series exhibit a roughly periodic signal
that resembles the scan pattern in Fig.~\ref{fig:bolos_mix}b. While
these correlated signals also clearly appear at both wavelengths,
referring to the maps of eigenvalues, they are considerably stronger
at 850\,\micron\ (consistent with the visual appearance of the
bolometer signals in Fig.~\ref{fig:pca}). Also note that the PSDs for
these eigenvectors exhibit nearly a pure $1/f$ signature with almost
no line features, or a white-noise plateau (suggesting that these are
high-\snr\ measurements of a purely low-frequency drift).

\begin{figure*}
\centering
\includegraphics[width=0.85\linewidth]{pca}
\caption{The first nine components from a principal component
  analysis, ranked by the mean eigenvalues of combined 450 (s4a
  subarray) and 850\,\micron\ (s8b subarray) time-series bolometer
  data, for the same observation as that used in
  Figs.~\ref{fig:bolos_mix}b and \ref{fig:pspec}b. For each component
  the top plot shows the time-series of the eigenvectors normalised by
  their \rms, the middle plot its PSD, and the bottom coloured panels
  the eigenvalues for the bolometers at each wavelength (the
  amplitudes of the eigenvectors in the time-series). For reference,
  both the mean, $\bar{\lambda}$, and \rms, $\lambda_\mathrm{rms}$,
  eigenvalues for the bolometers in each subarray are also shown. Most
  of the correlated $1/f$ signal is encapsulated in Component~1, and
  appears to be dominated by scan-synchronous noise (compare with the
  scan pattern in Fig.~\ref{fig:bolos_mix}b). Some of this signal also
  appears in Component~3. Component~2 is another $1/f$ signal that is
  dominant in the 850\,\micron\ subarray. Most of the remaining
  components consist of weaker $1/f$ noise and line features
  predominantly above $\sim$30\,Hz. The eigenvalue maps show
  correlated patterns along columns suggesting pickup in their common
  amplifier chains.}
\label{fig:pca}
\end{figure*}

@@

The approach taken by SMURF to reducing \scuba\ data is to model (and
remove) predominantly low-frequency noise sources that are correlated
amongst detectors, and to iterate this process along with estimates of
the map. Significant experimentation with different models for noise
sources during \scuba\ commissioning required a highly flexible
software framework, and a configurable user interface. To achieve this
goal, while minimizing development time, it was decided to build SMURF
as a Starlink package \citep{2009ASPC..411..418J}, which provides
access to a large suite of libraries (including a commanding and
messaging interface, astrometric coordinate conversions and generation
of standard WCS information, file formats etc.). Furthermore,
Starlink\footnote{\url{http://www.starlink.ac.uk}} is
open-source\footnote{\url{https://github.com/Starlink}} (distributed
under the GNU General Public Licence v3), and already used extensively
at the Joint Astronomy Centre (host of the JCMT) for many other
systems \citep{jenness2011}, which helps with interoperability. Though
originally written in FORTRAN, many of the core Starlink libraries are
now ported to native C, or at least have a C interface. It was
therefore decided to develop SMURF in C as well (rather than C++, for
example, which would have required adding further dependencies to
Starlink), although we have taken an object-oriented approach. For
example, all of the data for a given signal component model are
encapsulated in a C structure, and there is a standard interface for
all functions that handle models (member data and functions for the
class, respectively). In this way it is easy to extend SMURF with new
models. Parallelisation is incorporated in the most time-consuming
low-level routines using threads (e.g., performing Fast Fourier
Transforms, re-gridding the data), usually handling either independent
blocks of bolometers, or blocks of time, in each thread, depending on
the nature of the calculation. We have found that for typical data
sets the processing time scales well with the number of central
processing unit (CPU) cores, although beyond 8 the returns are
diminished due to tasks that cannot be parallelised (e.g., reading the
data from disk).

@@

\begin{figure*}
\raggedright
\textbf{(a) large step} \hspace{7.3cm} \textbf{(b) small steps}\\

\vspace{0.2cm}

\includegraphics[width=\linewidth]{steps1}
\caption{Examples of steps in time-series data. Steps occur with a
wide range of heights from the large steps shown in (a) to the small
steps shown in (b). Large steps are often followed by a brief
``over-shoot'' as shown in (a) which is ringing caused by the
anti-aliasing filter. In both plots, the black curve is the
uncorrected time-series, and the red curve is the corrected time
series. Samples close to a step are omitted in the corrected
time-series.  In (b) the blue curve is the uncorrected time-series for
a nearby bolometer. The similarity between the red and blue curves
shows that the step correction is performing well.  }
\label{fig:steps1}
\end{figure*}

@@

Steps occur with a wide range of heights and shapes. The ratio of step
height to noise can vary from less than 10 to several hundred. Some
steps occur over a single sample, such as the step close to sample
5000 in Fig.~\ref{fig:steps1}b, but others happen more gradually, such
as the step close to sample 5300. In addition, a step can be preceded
or followed by a short period of instability, as is visible at the
bottom of the step in Fig.~\ref{fig:steps1}a (this is probably due to
the response of the \scuba\ anti-aliasing filter; the sudden large
step occurs prior to the 200\,Hz re-sampling). Further problems are
caused by steps that occur close together in time, such as the large
downward step followed by a smaller upward step close to sample 5000
in Fig.~\ref{fig:steps1}b.

@@

The primary purpose of the noise component, \model{NOI}, is to measure
the white noise levels of each bolometer, which is approximated with a
single (non-time varying) variance, $\sigma_i^{\mathbf{w}2}$. This
value is then used to calculate weights for map estimation,
$\mathbf{w}_i = 1/\sigma_i^{\mathbf{w}2}$. The measurement generally
occurs once all of the other models have been fit and removed [i.e.,
once $\mathbf{n}^\mathrm{w}_i(t)$ from Eq.~\ref{eq:noise} is
isolated].

@@

The map-maker will halt after a user-specified number of iterations,
or once some convergence criterion has been achieved. Presently two
numerical quantities are tracked after each iteration: the change in
reduced chi-squared, $\chi^2_\mathrm{r}$; and the change in the
map. While convergence in $\chi^2_\mathrm{r}$ is not, ultimately, a
good indicator of when the solution should be stopped, we discuss its
calculation and properties first as an introduction to the problem of
model degeneracies.

The bolometer residual variances, $\sigma^2_i$, are measured once the
modelled signal components have been removed. In the early iterations,
the residuals contain both white noise, and other long-timescale
features. However, when the solution has converged, this signal should
look white. Taking the component of bolometer variances due to white
noise, $\sigma_i^{\mathbf{w}2}$ (measured from the bolometer PSDs in
\model{NOI}, Section~\ref{sec:noi}), $\chi^2_\mathrm{r}$ is calculated
as $(1/N) \sum_i \sigma^2_i / \sigma_i^{\mathbf{w}2}$, where $i$ runs
over the $N$ bolometers. In other words, it is the average ratio
between the measured time-series bolometer variances and their
white-noise levels measured between 2 and 10\,Hz, and should tend to a
value of $\sim$1 if the low-frequency noise (and any bright
astronomical signals) have been successfully removed. However, this
expression does not account for the degrees of freedom in the
model. Clearly there are a large number of parameters that will
``fit-out'' some of the uncorrelated white noise, and bias this
estimate of $\chi^2_\mathrm{r}$ low. Indeed, for a converged map,
values near $\sim$0.85 are common, depending primarily on the fraction
of the total bandwidth that has been removed by high-pass filtering
(Section~\ref{sec:flt}). Regardless of the normalisation, once the
model parameterisation has been chosen, this quantity should converge
to some fixed value.

Practically speaking, $\chi^2_\mathrm{r}$ tends to converge before the
map due to model degeneracies. One simple example is the degeneracy
between large-scale astronomical structures (larger than the array
footprint), and the common-mode rejection step (\model{COM}), which
will be illustrated in Section~\ref{sec:point}. In this case, there is
a significant space of maps with spurious large-scale structures that
are anti-correlated with features in the common-mode, and the solution
is free to wander this space while producing flat residuals.
